MTBD models with partner notification

### 2023-07-25 article/article.tex
Simplified BD-Pn equation solutions

Phylodynamic models can be classified into two main families:  coalescent~\cite{Volz2009a,Drummond2005,Pybus2000a} and birth-death (BD)~\cite{Kendall1948,Maddison2007,Stadler2009,Stadler2010}. Coalescent models are often preferred for estimating deterministic population dynamics, however for highly stochastic processes, such as the dynamics of emerging pathogens, BD models are better adapted~\cite{Macpherson2021}. In the classic BD model with incomplete sampling (BDS~\cite{Stadler2009}), births represent pathogen transmission events (happening at a constant transmission rate), while deaths correspond to becoming non-infectious (e.g. due to healing, self-isolation, starting a treatment, or death, modelled with a constant removal rate).

@@

In this study we propose an extension of the MTBD framework that allows to model non-random sampling due to partner notification, MTBD-PN. In the rest of the article, we describe the MTBD-PN model family and its assumptions, implement a transmission tree simulator under these models, derive the equations for their likelihood calculation, implement a maximum likelihood parameter estimator and test it on simulated data. We also propose a test allowing to distinguish between MTBD and MTBD-PN models. Finally, we apply it to the study of HIV epidemic in the UK.

@@

\begin{equation}
\scriptsize
\begin{cases}
\dot{U}_{s}(t) = &\Big(\sum\limits_{k=1}^{m} \mu_{sk} + \sum\limits_{k=1}^{m} \lambda_{sk} + \psi_s\Big) U_{s}(t)\; \textit{\color{gray} $\leftarrow$ no event in the next infinitesimal time $\Delta t$ }\\ 
    & - \sum\limits_{k=1}^{m} \mu_{sk} U_{k}(t)\;  \textit{\color{gray} $\leftarrow$ change of the state, folowed by unsampled evolution from the new state }\\
    &- \sum\limits_{k=1}^{m} \lambda_{sk} U_{s}(t)U_{k}(t) \;  \textit{\color{gray} $\leftarrow$ transmission, followed by unsampled evolutions of both subtrees}\\
    &- \psi_s (1 - \rho)\;  \textit{\color{gray} $\leftarrow$ removal without sampling}\\
U_{s}(T) = & 1\;  \textit{\color{gray} $\leftarrow$ the probability to stay unsampled over time 0 is 1} \label{eq:u}
\end{cases}
\end{equation}

With the equations~(\ref{eq:u}), we can write down the equations for the probability densities $p_{s,si}^{(i)}(t)$ of evolving as on an observed tree branch that finishes at time $t_i$ in state $si$, starting  on this branch at a time $t \leq t_i$ in state $s$. The master equations for the MTBD models were initially developed by ~\citet{Stadler2013a}, however here we present their parallelizable version that we recently proposed~\cite{zhukovaFastAccurateMaximumLikelihood2022}:

@@

If the states of all the observed tree $\mathscr{T}$ nodes are known, then its likelihood density under parameter values $\Theta$ can be calculated as $L(\mathscr{T}|\Theta)$:

@@

We now propose a non-Markovian extension of the MTBD framework that adds partner notification (PN), for a tree with known node states.  In MTBD-PN, at the moment of sampling the sampled individual might notify their most recent partner with a given probability $\rho_n$. Upon notification, partner is removed almost instantaneously (modelled via a notified removal rate $\psi_n >> \psi_k \;\forall k \in 1:m$). MTBD-PN hence has $2k^2 + 1$ classical MTBD and 2 additional parameters:
\begin{itemize}
 \item $\mu_{kl} \;\forall k,l \in 1:m, k \neq l$ -- state change rate from $k$ to $l$;
 \item $\lambda_{kl} \;\forall k,l \in 1:m$ -- transmission rate: an infected individual in state $k$ transmits their pathogen (which corresponds to an internal node in the transmission tree), the newly infected individual is in state $l$;
 \item $\psi_k \;\forall k \in 1:m$ -- removal rate: an infected individual in state $k$ stops being infectious at this rate (exponential distribution), which corresponds to a tip in the transmission tree (observed or hidden);
 \item $\rho$ -- sampling probability: upon removal the individual's pathogen might get sampled with this probability, which corresponds to an observed tip in the transmission tree;
 \item $\rho_n$ -- notification probability: upon sampling, the sampled individual might notify their most recent partner with this probability;
 \item $\psi_n >> \psi_k  \;\forall k \in 1:m$ -- notified removal rate: a notified individual stops being infectious and gets observed with this rate (exponential distribution), which corresponds to an observed tip in the transmission tree.
\end{itemize}

This model is a simplification of the notification process, in particular it makes 5 assumptions, which might need to be relaxed in the future work:
\begin{enumerate}
\item only observed individuals can notify (instead of any removed individual);
\item notified individuals are always observed upon removal;
\item notified individuals do not notify further;
\item after the notification notified individuals do not transmit further;
\item only the most recent partner can get notified.
\end{enumerate}

These assumptions make the mathematics behind the model manageable, however they can be easily relaxed in a simulator (e.g. to be used with a deep learning estimator~\cite{Voznica2021} after).

\subsection{Mathematical formulation}
$p_{s,si}^{(i)}(t)$ in equations~(\ref{eq:p}) describes an evolution along a tree branch, provided that we do not know whether the individual at the end of the branch (at time $t_i$) is the same as at time $t$. They could be different if a hidden transmission, where the donor subtree stayed unobserved, occurred between times $t$ and $t_i$.

We will use an \textbf{oriented probability} $P^{(i,o)}_{s,si}(t)$ for the cases when the individual at the end of the branch is the same as at time $t$:

\begin{equation}
\scriptsize
\begin{cases}
\dot{p}_{s,si}^{(i,o)}(t) = & \Big(\sum\limits_{k=1}^{m} \mu_{sk} + \sum\limits_{k=1}^{m} \lambda_{sk} + \psi_s\Big) p_{s,si}^{(i,o)}(t)\; \textit{\color{gray} $\leftarrow$ no event in the next infinitesimal time $\Delta t$ }\\
    & - \sum\limits_{k=1}^{m} \mu_{sk} p_{k,si}^{(i,o)}(t)\;  \textit{\color{gray} $\leftarrow$ change of the state, folowed by evolution from the new state }\\
    & - \sum\limits_{k=1}^{m} \lambda_{sk} p_{s,si}^{(i,o)}(t)U_{k}(t)\;  \textit{\color{gray} $\leftarrow$ transmission, where the recipient subtree stayed unsampled}\\
p_{s,si}^{(i,o)}(t_i) = & 
    \begin{cases}
    0 & \textit{\color{gray} if $i$ is in state $k \neq si$}\\
    1 & \textit{\color{gray} if $i$ is in state $si$}
    \end{cases}\label{eq:p-o}
\end{cases}
\end{equation}

@@

\begin{equation}
\scriptsize
\begin{cases}
\dot{p}_{s,si}^{(i,nh)}(t) = & \Big(\sum\limits_{k=1}^{m} \mu_{sk} + \sum\limits_{k=1}^{m} \lambda_{sk} + \psi_s\Big) p_{s,si}^{(i,nh)}(t)\; \textit{\color{gray} $\leftarrow$ no event in the next infinitesimal time $\Delta t$ }\\
    & - \sum\limits_{k=1}^{m} \mu_{sk} p_{k,si}^{(i,nh)}(t)\;  \textit{\color{gray} $\leftarrow$ change of the state, folowed by evolution from the new state }\\
p_{s,si}^{(i,nh)}(t_i) = & 
    \begin{cases}
    0 & \textit{\color{gray} if $i$ is in state $k \neq si$}\\
    1 & \textit{\color{gray} if $i$ is in state $si$}
    \end{cases}\label{eq:p-nh}
\end{cases}
\end{equation}

Under the MTBD-PN model, the probability density $q_{si,sj}(i)$ of a branch connecting a node $i$ in state $si$ to its parent node $j$ in state $sj$ can be written as:

\begin{equation}
\scriptsize
q_{sj,si}(i) = 
\begin{cases}
q^{(u)}_{sj,si}(i) = p^{(i)}_{sj,si}(t_j) \textit{\color{gray} ~~if $i$ is an \textbf{unnotified non-notifier}}\\
q_{sj,si}^{(n)}(i) = 
p_{sj,si}^{(i,nh)}(t_j) \textit{\color{gray} ~~if $i$ is a tip corresponding to a \textbf{notifier}}\\
~~~~~~~~~~~~ \textit{\color{gray}(parent is the last transmission, no hidden transmission along the branch)}
\\
q^{(p)}_{sj,si}(i)  = p^{(i, o)}_{sj,si}(t_j)  \textit{\color{gray} ~~if $i$ is an internal node who will be notified after $t_i$}
\\
q^{(p)}_{sj,si}(i, r)  = p^{(r, o)}_{sj,si}(t_j) e^{- \psi_n (t_i - t_r)} \textit{\color{gray} if $i$ is a \textbf{partner} tip notified by $r$ at time $t_r < t_i$}\\
\end{cases}
\label{eq:branch}
\end{equation}

Using the formulas~(\ref{eq:branch}), we can calculate the likelihood density of a tree using a pruning algorithm~\cite{10.1093/sysbio/22.3.240}, where for each visited node $i$ we calculate two values for each possible state $sj \in 1:m$ at the beginning of the branch connecting it to its parent $j$: the likelihood density of its subtree if its branch corresponds to (1) an unnotified individual $l^{(u)}_{sj}(i)$; (2) a notifier $l^{(u)}_{sj}(i)$ (for tips only); or (3) an (eventually) notified partner $l^{(p)}_{sj}(i, r)$ (notified by $r$).

For a tip $i$, these values can be calculated by considering all its possible states $si \in 1:m$, as:
\begin{equation}
\scriptsize
\begin{array}{ll}
l^{(u)}_{sj}(i) &= \sum\limits_{si=1}^{m} q^{(u)}_{sj,si}(i)\psi_{si}\rho (1-\rho_{n})\\
l^{(n)}_{sj}(i) &= \sum\limits_{si=1}^{m} q^{(n)}_{sj,si}(i)\psi_{si}\rho \rho_{n}\\
l^{(p)}_{sj}(i, r) &= \sum\limits_{si=1}^{m} q^{(p)}_{sj,si}(i,r)\psi_{n}\\
\end{array}
\label{eq:tip}
\end{equation}

@@

\begin{equation}
\scriptsize
\begin{split}
l^{(u)}_{sj}(i) = &\sum\limits_{si=1}^{m}q^{(u)}_{sj,si}(i) \sum\limits_{k=1}^{m} \lambda_{si,k} \\
\cdot \Big(~&
l^{(u)}_{si}(i0)l^{(u)}_{k}(i1) + l^{(u)}_{k}(i0)l^{(u)}_{si}(i1)  \textit{\color{gray} $\leftarrow$ no notification on child branches} \\
& + is\_tip(i0) \big(l^{(n)}_{si}(i0)l^{(p)}_{k}(i1,i0) + l^{(n)}_{k}(i0)l^{(p)}_{si}(i1,i0) \big)\textit{\color{gray} $\leftarrow$ notification of someone in $i1$'s subtree by $i0$} \\
&+ is\_tip(i1) \big(l^{(p)}_{si}(i0,i1)l^{(n)}_{k}(i1) + l^{(p)}_{k}(i0,i1)l^{(n)}_{si}(i1) \big)~\Big) \textit{\color{gray} $\leftarrow$ notification of someone in $i0$'s subtree by $i1$},\\
\textit{where } & is\_tip(r) = 1 \textit{ if r is a tip and $= 0$ if r is an internal node.}
\end{split}\label{eq:lu}
\end{equation}

Note that a notification will only be possible if (1) the notifying branch is external (as the notifier only notifies the most recent partner); (2) there is at least one tip in the notified subtree that can be a partner, i.e. was sampled after the notifier and whose parent's time was before notification (i.e. there was no transmission between notification and sampling).

@@

\begin{equation}
\scriptsize
\begin{split}
l^{(p)}_{sj}(i, r) = \sum\limits_{si=1}^{m}q^{(p)}_{sj,si}(i,r) \sum\limits_{k=1}^{m}  \lambda_{si,k}\Big(&l^{(p)}_{si}(i0,r)l^{(u)}_{k}(i1) \textit{\color{gray} $\leftarrow$ notification in $i0$'s subtree} \\
 &+ l^{(u)}_{k}(i0)l^{(p)}_{si}(i1,r)\Big)  \textit{\color{gray} $\leftarrow$ notification in $i1$'s subtree} 
\end{split}\label{eq:lu}
\end{equation}

Note that $l^{(p)}_{sj}(i, r)$ depends on the notifier $r$ and can only be calculated once the notification time is known, i.e. once we have climbed to an internal node corresponding to the transmission between the notifier (corresponding to one of its children, must be a tip) and the partner (the other child). As only the contributions coming from the partner tip branches depend on the notification times, while climbing the tree for each internal node $i$ and each of the possible states $sj \in 1:m$ at the beginning of its branch, we can store an array of pre-calculated values for notified internal paths for each of its tips (e.g. $a$) and each of the states $v \in 1:m$ at the beginning of $a$'s branch, describing the tip's sampling time $t_a$ and a configuration $C_{sj}(i,a,v)$:

\begin{equation}
\scriptsize
\begin{split}
C_{sj}(i,a,v) = &\begin{cases}
q^{(p)}_{sj,v}(i)
\sum\limits_{k=1}^{m}\lambda{v,k} \cdot l_{k}^{(u)}\Big(child(i, \not a)\Big) \textit{\color{gray}~~~if $i$ is a direct parent of $a$}
\\
\sum\limits_{si=1}^{m}q^{(p)}_{sj,si}(i)
\sum\limits_{k=1}^{m}\lambda{si,k} \cdot l_{k}^{(u)}\Big(child(i, \not a)\Big)C_{si}\Big(child(i, a),a,v\Big) \textit{\color{gray}~~otherwise}\\ 
\end{cases},
\\ &\textit{where child(i,a) is the child of i whose subtree contains a},
\\
 &\textit{~~~~~~~~child(i,$\not a$) is the child of i whose subtree does not contain a.}
\end{split}
\label{eq:c}
\end{equation}

@@

Once a potential notifier tip $r$ is identified (i.e., $r$ is a sister tip of the currently visited node $i$), we can calculate $l^{(p)}_{sj}(i, r)$ as:
\begin{equation}
\scriptsize 
l^{(p)}_{sj}(i, r) = \sum\limits_{a \in tips(i)}\sum\limits_{k=1}^{m} C_{sj}(i, a, k) l^{(p)}_{k}(a, r)\label{eq:lp}
\end{equation}

@@

\subsection{BD-PN}
A particular example of the MTBD model generating trees with known node states is the basic birth-death model~\citep{Stadler2009} where all the individuals are in the same state. They can transmit with a constant rate $\lambda$, get removed with a constant rate $\psi$, be sampled upon removal with a probability $\rho$. Below we describe the special case of equations~(\ref{eq:u}-\ref{eq:lp}) for its PN extension BD-PN.

@@

\begin{equation}
\scriptsize
\begin{cases}
\dot{U}(t) &= \Big(\lambda + \psi\Big) U(t) - \lambda U^2(t) - \psi (1 - \rho)\\\\
U(T) &=  1\\
\dot{p}^{(i)}(t) &=  \Big(\lambda + \psi\Big) p^{(i)}(t) - 2 \lambda p^{(i)}(t)U(t)\\
\dot{p}^{(i,o)}(t) &=  \Big(\lambda + \psi\Big) p^{(i)}(t) - \lambda p^{(i)}(t)U(t)\\
\dot{p}^{(i,nh)}(t) &=  \Big(\lambda + \psi\Big) p^{(i)}(t)\\
p^{(i)}(t_i) &= p^{(i,o)}(t_i) = p^{(i,nh)}(t_i) = 1
\end{cases}
\end{equation}

These equations have the following closed form solutions:
\begin{equation}
\scriptsize
\begin{split}
&\begin{cases}
U(t) = \frac{\lambda + \psi + c_1 \frac{c_2 e^{t c_1} - 1}{c_2 e^{tc_1} + 1}}{2\lambda}\\
\\
p^{(i)}(t) = \frac{c_3^2 e^{c_1 (t - t_i)}}{(c_2e^{c_1 t} + 1)^2} \\
\\
p^{(i,o)}(t) =  \frac{c_3 e^{1/2 (\lambda + \psi + c_1)(t - t_i)}}{c_2e^{c_1 t} + 1} \\
\\
p^{(i,nh)}(t) =  e^{(\lambda + \psi)(t - t_i)}
\end{cases},\\
& \textit{where } c_1 = \sqrt{(\lambda - \psi)^2 + 4 \lambda\psi\rho},\; c_2 = \frac{c_1 + \lambda - \psi}{c_1 - \lambda + \psi}e^{-Tc_1},\; c_3 = c_2 e^{c_1 t_i} + 1
\end{split}
\end{equation}

The likelihood of a tree $\mathscr{T}$ under BD-PN model with parameter values $\Theta$ can be calculated as $\mathscr{L}(\mathscr{T}|\Theta) = l^{(u)}(root)$, where $l^{(u)}(i)$ is a likelihood of node $i$'s subtree if the branch connecting $i$ to its parent node (of length zero for the root) correspond to an unnotified individual. It can be calculated, along with $l^{(p)}(i, r)$ (likelihood when $i$'s branch corresponds to an eventually notified partner, notified by $r$) for each tree node $i$ with a pruning algorithm.

>>>>

Phylodynamic models can be classified into two main families:  coalescent~\cite{Volz2009a,Drummond2005,Pybus2000a} and birth-death (BD)~\cite{Kendall1948,Maddison2007,Stadler2009,Stadler2010}. Coalescent models are often preferred for estimating deterministic population dynamics, however for highly stochastic processes, such as the dynamics of emerging pathogens, BD models are better adapted~\cite{Macpherson2021}. In the classic BD model with incomplete sampling~\cite{Stadler2009}, births represent pathogen transmission events (happening at a constant transmission rate), while deaths correspond to becoming non-infectious (e.g. due to healing, self-isolation, starting a treatment, or death, modelled with a constant removal rate).

@@

In this study we propose an extension of the MTBD framework that allows for modeling non-random sampling due to partner notification (PN), MTBD-PN. In the rest of the article, we describe the MTBD-PN model family and its assumptions, present a transmission tree simulator under these models, derive the equations for MTBD-PN likelihood calculation, and propose a test for detecting partner notification in pathogen phylogenetic trees. We also present a maximum likelihood parameter estimator for the simplest model of the MTBD-PN family, BD-PN, and test it on simulated data. Finally, we apply it to the study of HIV-1 B epidemic in the UK.

@@

\begin{equation}
\scriptsize
\begin{cases}
\dot{U}_{s}(t) = &\Big(\sum\limits_{k=1}^{m} \mu_{sk} + \sum\limits_{k=1}^{m} \lambda_{sk} + \psi_s\Big) U_{s}(t)\; \textit{\color{gray} $\leftarrow$ no event in the next infinitesimal time $\Delta t$ }\\ 
    & - \sum\limits_{k=1}^{m} \mu_{sk} U_{k}(t)\;  \textit{\color{gray} $\leftarrow$ change of the state, folowed by unsampled evolution from the new state }\\
    &- \sum\limits_{k=1}^{m} \lambda_{sk} U_{s}(t)U_{k}(t) \;  \textit{\color{gray} $\leftarrow$ transmission, followed by unsampled evolutions of both subtrees}\\
    &- \psi_s (1 - \rho)\;  \textit{\color{gray} $\leftarrow$ removal without sampling}\\
U_{s}(T) = & 1\;  \textit{\color{gray} $\leftarrow$ the probability to stay unsampled over time 0 is 1} \label{eq:Us}
\end{cases}
\end{equation}

With the equations~(\ref{eq:Us}), we can write down the equations for the probability densities $p_{s,si}^{(i)}(t)$ of evolving as on an observed tree branch that finishes at time $t_i$ in state $si$, starting  on this branch at a time $t \leq t_i$ in state $s$. The master equations for the MTBD models were initially developed by ~\citet{Stadler2013a}, however what we present here, is their parallelizable version that we proposed in~\cite{zhukovaFastAccurateMaximumLikelihood2022}:

@@

If the states of all the observed nodes of the tree $\mathscr{T}$ are known, then its likelihood density under parameter values $\Theta$ can be calculated as $L(\mathscr{T}|\Theta)$:

@@

We now propose a non-Markovian extension of the MTBD framework that adds partner notification (PN), for a tree with known node states.  In MTBD-PN, at the moment of sampling the sampled individual might notify their most recent partner with a given probability $\rho_n$. Upon notification, partner is removed almost instantaneously (modeled via a notified removal rate $\psi_n >> \psi_k \;\forall k \in 1:m$). MTBD-PN hence has $2k^2 + 1$ classical MTBD and 2 additional parameters:
\begin{itemize}
 \item $\mu_{kl} \;\forall 1 \leq k,l  \leq m, k \neq l$ -- constant state change rate from $k$ to $l$;
 \item $\lambda_{kl} \;\forall 1 \leq k,l \leq m$ -- constant transmission rate: an infected individual in state $k$ transmits their pathogen (which corresponds to an internal node in the transmission tree), the newly infected individual is in state $l$;
 \item $\psi_k \;\forall 1 \leq k \leq m$ -- constant removal rate: an infected individual in state $k$ stops being infectious at this rate. This corresponds to a tip in the transmission tree (observed or hidden);
 \item $\rho$ -- constant sampling probability: upon removal the individual's pathogen might get sampled with this probability, which corresponds to an observed tip in the transmission tree;
 \item $\rho_n$ -- constant notification probability: upon sampling, the sampled individual might notify their most recent partner with this probability;
 \item $\psi_n >> \psi_k  \;\forall 1 \leq k \leq m$ -- constant notified removal rate: a notified individual (if not yet removed via standard procedure) stops being infectious and gets observed with this rate. This creates an observed tip in the transmission tree.
\end{itemize}

This model is a simplification of the notification process, in particular it makes 5 assumptions:
\begin{enumerate}
\item only observed individuals can notify (instead of any removed individual);
\item notified individuals are always observed upon removal;
\item notified individuals do not notify further;
\item after the notification, notified individuals do not transmit further;
\item only the most recent partner can get notified.
\end{enumerate}

These assumptions make the mathematics behind the model manageable, however they can be easily relaxed in a tree simulator (e.g., to be used with a deep learning estimator~\cite{Voznica2021}, which does not require likelihood calculation).

\subsubsection{Mathematical formulation}
$p_{s,si}^{(i)}(t)$ in equations~(\ref{eq:p}) describes an evolution along a tree branch, provided that we do not know whether the individual at the end of the branch (at time $t_i$) is the same as at time $t$. These two individuals could be different if a hidden transmission, where the donor subtree stayed unobserved, occurred between times $t$ and $t_i$.

We will use an \textbf{oriented probability} $p^{(i,o)}_{s,si}(t)$ for the cases when the individual at the end of the branch is the same as at time $t$:

\begin{equation}
\scriptsize
\begin{cases}
\dot{p}_{s,si}^{(i,o)}(t) = & \Big(\sum\limits_{k=1}^{m} \mu_{sk} + \sum\limits_{k=1}^{m} \lambda_{sk} + \psi_s\Big) p_{s,si}^{(i,o)}(t)\; \textit{\color{gray} $\leftarrow$ no event in the next infinitesimal time $\Delta t$ }\\
    & - \sum\limits_{k=1}^{m} \mu_{sk} p_{k,si}^{(i,o)}(t)\;  \textit{\color{gray} $\leftarrow$ change of the state, followed by evolution from the new state }\\
    & - \sum\limits_{k=1}^{m} \lambda_{sk} p_{s,si}^{(i,o)}(t)U_{k}(t)\;  \textit{\color{gray} $\leftarrow$ transmission, where the recipient subtree stayed unsampled}\\
p_{s,si}^{(i,o)}(t_i) = & 
    \begin{cases}
    0 & \textit{\color{gray} if $i$ is in state $k \neq si$}\\
    1 & \textit{\color{gray} if $i$ is in state $si$}
    \end{cases}\label{eq:p-o}
\end{cases}
\end{equation}

@@

\begin{equation}
\scriptsize
\begin{cases}
\dot{p}_{s,si}^{(i,nh)}(t) = & \Big(\sum\limits_{k=1}^{m} \mu_{sk} + \sum\limits_{k=1}^{m} \lambda_{sk} + \psi_s\Big) p_{s,si}^{(i,nh)}(t)\; \textit{\color{gray} $\leftarrow$ no event in the next infinitesimal time $\Delta t$ }\\
    & - \sum\limits_{k=1}^{m} \mu_{sk} p_{k,si}^{(i,nh)}(t)\;  \textit{\color{gray} $\leftarrow$ change of the state, followed by evolution from the new state }\\
p_{s,si}^{(i,nh)}(t_i) = & 
    \begin{cases}
    0 & \textit{\color{gray} if $i$ is in state $k \neq si$}\\
    1 & \textit{\color{gray} if $i$ is in state $si$}
    \end{cases}\label{eq:p-nh}
\end{cases}
\end{equation}

Under the MTBD-PN model, the probability density $q_{si,sj}(i)$ of a branch connecting a node $i$ in state $si$ to its parent node $j$ in state $sj$ (without taking into account the event at the end of the branch) can be written as:

\begin{equation}
\scriptsize
q_{sj,si}(i) = 
\begin{cases}
q^{(u)}_{sj,si}(i) = p^{(i)}_{sj,si}(t_j) \textit{\color{gray} ~if $i$ is an \textbf{unnotified non-notifier}}\\
q_{sj,si}^{(n)}(i) = 
p_{sj,si}^{(i,nh)}(t_j) \textit{\color{gray} ~~if $i$ is a tip corresponding to a \textbf{notifier}}\\
~~~~~~~~~~~~ \textit{\color{gray}(parent is the last transmission, no hidden transmission along the branch)}
\\
q^{(p)}_{sj,si}(i)  = p^{(i, o)}_{sj,si}(t_j)  \textit{\color{gray} ~if $i$ is an internal node whose individual will be notified after $t_i$}
\\
q^{(p)}_{sj,si}(i, r)  = 
\begin{cases}
p^{(r, o)}_{sj,si}(t_j) e^{- \psi_n (t_i - t_r)} \textit{\color{gray} if $i$ is a \textbf{partner} tip notified by $r$ at time $t_r \leq t_i$}\\
p^{(i, o)}_{sj,si}(t_j) \textit{\color{gray}~~ if $i$ is a \textbf{partner} tip sampled via a standard procedure} \\
~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~\textit{\color{gray} before the notification by $r$: $t_i < t_r$}\\
\end{cases}
\end{cases}
\label{eq:branch}
\end{equation}

Using the formulas~(\ref{eq:branch}), we can calculate the likelihood density of a tree using a pruning algorithm~\cite{10.1093/sysbio/22.3.240}, where for each visited node $i$, and for each possible state $1 \leq sj \leq m$ at the beginning of the branch connecting it to its parent $j$  we calculate the likelihood density of its subtree if its branch corresponds to (1) an unnotified individual $l^{(u)}_{sj}(i)$; (2) a notifier $l^{(n)}_{sj}(i)$ (for tips only); or (3) an (eventually) notified partner $l^{(p)}_{sj}(i, r)$ (notified by the individual corresponding to the tip $r$ at time $t_r$).

For a tip $i$ in state $si$ these values can be calculated as:
\begin{equation}
\scriptsize
\begin{array}{ll}
l^{(u)}_{sj}(i) &= q^{(u)}_{sj}(i)\psi_{si}\rho (1-\rho_{n})\\
l^{(n)}_{sj}(i) &= q^{(n)}_{sj}(i)\psi_{si}\rho \rho_{n}\\
l^{(p)}_{sj}(i, r) &= q^{(p)}_{sj}(i,r)\psi_{n}\\
\end{array}
\label{eq:tip}
\end{equation}

@@

\begin{equation}
\scriptsize
\begin{split}
l^{(u)}_{sj}(i) = &q^{(u)}_{sj}(i) \sum\limits_{k=1}^{m} \lambda_{si,k} \\
\cdot \Big(~&
l^{(u)}_{si}(i0)l^{(u)}_{k}(i1) + l^{(u)}_{k}(i0)l^{(u)}_{si}(i1)  \textit{\color{gray} $\leftarrow$ no notification on child branches} \\
& + is\_tip(i0) \big(l^{(n)}_{si}(i0)l^{(p)}_{k}(i1,i0) + l^{(n)}_{k}(i0)l^{(p)}_{si}(i1,i0) \big)\textit{\color{gray} $\leftarrow$ notification of someone in $i1$'s subtree by $i0$} \\
&+ is\_tip(i1) \big(l^{(p)}_{si}(i0,i1)l^{(n)}_{k}(i1) + l^{(p)}_{k}(i0,i1)l^{(n)}_{si}(i1) \big)~\Big) \textit{\color{gray} $\leftarrow$ notification of someone in $i0$'s subtree by $i1$},\\
\textit{where } & is\_tip(r) = 1 \textit{ if r is a tip and $= 0$ if r is an internal node.}
\end{split}\label{eq:lu}
\end{equation}

Note that a notification will only be possible if the notifying branch is external as the notifier only notifies the most recent partner, i.e. the transmission corresponds to the notifier's parent node. %; (2) there is at least one tip in the notified subtree that can be a partner, i.e. was sampled after the notifier and whose parent's time was before notification (i.e. there was no transmission between notification and sampling).

@@

\begin{equation}
\scriptsize
\begin{split}
l^{(p)}_{sj}(i, r) = q^{(p)}_{sj}(i,r) \sum\limits_{k=1}^{m}  \lambda_{si,k}\Big(&l^{(p)}_{si}(i0,r)l^{(u)}_{k}(i1) \textit{\color{gray} $\leftarrow$ notification in $i0$'s subtree} \\
 &+ l^{(u)}_{k}(i0)l^{(p)}_{si}(i1,r)\Big)  \textit{\color{gray} $\leftarrow$ notification in $i1$'s subtree} 
\end{split}\label{eq:lu}
\end{equation}

Note that $l^{(p)}_{sj}(i, r)$ depends on the notifier $r$ and can only be calculated once the notification time is known, i.e. once we have climbed to an internal node corresponding to the transmission between the notifier (corresponding to one of its children, must be a tip) and the partner (the other child). In the partner subtree only the contribution coming from the partner tip branch depends on the notification time. Hence while climbing the tree from tips to the root, for each internal node $i$ and each of the possible states $sj \in 1:m$ at the beginning of its branch, we can store an array of pre-calculated values for notified internal paths leading to each of its tips. The values of this array will store for each tip (e.g. $a$) its sampling time $t_a$ and a configuration $C_{sj}(i,a)$, representing the subtree likelihood provided $a$ corresponds to the same notified partner individual as $i$ (but without the $a$'s tip branch contribution):

\begin{equation}
\scriptsize
\begin{split}
C_{sj}(i,a) = &\begin{cases}
1 \textit{\color{gray}~~~if $i = a$}
\\
q^{(p)}_{sj}(i)
\sum\limits_{k=1}^{m}\lambda{si,k} \cdot l_{k}^{(u)}\Big(child(i, \not a)\Big)C_{si}\Big(child(i, a),a\Big) \textit{\color{gray}~~otherwise}\\ 
\end{cases},
\\ &\textit{where child(i,a) is the child of i whose subtree contains a},
\\
 &\textit{~~~~~~~~child(i,$\not a$) is the child of i whose subtree does not contain a.}
\end{split}
\label{eq:c}
\end{equation}

@@

Once a potential notifier tip $r$ is identified (i.e., $r$ is a sister tip of the currently visited node $i$), we can calculate $l^{(p)}_{sj}(i, r)$ as:
\begin{equation}
\scriptsize 
l^{(p)}_{sj}(i, r) = \sum\limits_{a \in tips(i)}C_{sj}(i, a) l^{(p)}_{sa}(a, r)\label{eq:lp}
\end{equation}

@@

\subsection{BD-PN}
A particular example of the MTBD model generating trees with known node states is the basic birth-death model~\citep{Stadler2009} where all the individuals are in the same state. They can transmit with a constant rate $\lambda$, get removed with a constant rate $\psi$, be sampled upon removal with a probability $\rho$. Below we describe the special case of equations~(\ref{eq:Us}-\ref{eq:lp}) for its PN extension BD-PN.

%\begin{equation}
%\scriptsize
%\begin{cases}
%\dot{U}(t) &= \Big(\lambda + \psi\Big) U(t) - \lambda U^2(t) - \psi (1 - \rho)\\
%U(T) &=  1\\
%\dot{p}^{(i)}(t) &=  \Big(\lambda + \psi\Big) p^{(i)}(t) - 2 \lambda p^{(i)}(t)U(t)\\
%p^{(i)}(t_i) &= 1
%\end{cases}
%\end{equation}
%
%\begin{equation}
%L(\mathscr{T}|\{\lambda, \psi, \rho\}) = n log(\psi\rho) + \sum\limits_{i \in \substack{internal\\nodes}} 2\lambda p^{(i0)}(t_i) p^{(i1)}(t_i)
%\end{equation}
%

@@

\begin{equation}
\scriptsize
\begin{cases}
\dot{U}(t) &= \Big(\lambda + \psi\Big) U(t) - \lambda U^2(t) - \psi (1 - \rho)\\
U(T) &=  1\\
\dot{p}^{(i)}(t) &=  \Big(\lambda + \psi\Big) p^{(i)}(t) - 2 \lambda p^{(i)}(t)U(t)\\
\dot{p}^{(i,o)}(t) &=  \Big(\lambda + \psi\Big) p^{(i,o)}(t) - \lambda p^{(i,o)}(t)U(t)\\
\dot{p}^{(i,nh)}(t) &=  \Big(\lambda + \psi\Big) p^{(i,nh)}(t)\\
p^{(i)}(t_i) &= p^{(i,o)}(t_i) = p^{(i,nh)}(t_i) = 1
\end{cases}
\end{equation}

%
%These equations have the following closed form solutions:
%\begin{equation}
%\scriptsize
%\begin{split}
%&\begin{cases}
%U(t) = \frac{\lambda + \psi + c_1 \frac{c_2 e^{t c_1} - 1}{c_2 e^{tc_1} + 1}}{2\lambda}\\
%\\
%p^{(i)}(t) = \frac{c_3^2 e^{c_1 (t - t_i)}}{(c_2e^{c_1 t} + 1)^2} \\
%\\
%p^{(i,o)}(t) =  \frac{c_3 e^{1/2 (\lambda + \psi + c_1)(t - t_i)}}{c_2e^{c_1 t} + 1} \\
%\\
%p^{(i,nh)}(t) =  e^{(\lambda + \psi)(t - t_i)}
%\end{cases},\\
%& \textit{where } c_1 = \sqrt{(\lambda - \psi)^2 + 4 \lambda\psi\rho},\; c_2 = \frac{c_1 + \lambda - \psi}{c_1 - \lambda + \psi}e^{-Tc_1},\; c_3 = c_2 e^{c_1 t_i} + 1
%\end{split}
%\end{equation}
%
%
%These equations have the following closed form solutions:
%\begin{equation}
%\scriptsize
%\begin{split}
%&\begin{cases}
%U(t) = \frac{\lambda + \psi + c_1 \frac{c_2 e^{ c_1 (t-T)} - 1}{c_2 e^{c_1 (t-T)} + 1}}{2\lambda}\\
%\\
%p^{(i)}(t) = \Big(\frac{c_2 e^{c_1 (t_i - T)} + 1}{c_2e^{c_1 (t - T)} + 1}\Big)^2e^{c_1 (t - t_i)} \\
%\\
%p^{(i,o)}(t) =  \frac{(c_2 e^{c_1 (t_i -T)} + 1) \Big(e^{c_1 (t -t_i)} e^{(\lambda + \psi)(t - t_i)}\Big)^{1/2}}{c_2e^{c_1 (t - T)} + 1} \\
%\\
%p^{(i,nh)}(t) =  e^{(\lambda + \psi)(t - t_i)}
%\end{cases},\\
%& \textit{where } c_1 = \sqrt{(\lambda - \psi)^2 + 4 \lambda\psi\rho},\; c_2 = \frac{c_1 + \lambda - \psi}{c_1 - \lambda + \psi},
%\end{split}
%\end{equation}
